---
title: "dplyr 2.0.0 transition guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr 2.0.0 transition guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This transition guide is aimed at backend authors.
One of the big goals of dbplyr 2.0.0 is to ensure that all dbplyr generics can live in dplyr.
I've done my best to design than transition so you can write a backend that works with both dbplyr 1.0.0 and dbplyr 2.0.0, and the complete transition won't happen immediately but will play out over the next year or two.

(dbplyr 2.0.0 won't be released for at least 3 months after I notify all backend authors, and if you need more time I'm happy to extend this deadline.)

The basic idea is that no generics will "move"; instead dbplyr will add new generics and dplyr will eventually remove them.
This makes it straightforward to support both dbplyr 1.0.0 and 2.0.0 in a single package.

## DBI generics

When dbplyr was first created there was much spurious variation across DBI backends, and a layer of shims was need to make dbplyr usable.
Now thanks to the work of Kiril Muller and on others on DBITest and friends, these shims are no longer needed, and dbplyr will now call the DBI generics directly:

-   Transaction management (using inside `copy_to()` to make indexing as efficient as possible): `db_begin()` -\> `dbBegin()`, `db_commit()` -\> `dbCommit()`, and `db_rollback()` -\> `dbRollback()`
-   `copy_to()` now uses `dbWriteTable()` instead of `db_write_table()`. This also eliminates this eliminates `db_create_table()`, `db_insert_into()`, and `db_data_type()` that were used for initial creation, and `db_list_tables()`, `db_has_table()` and `db_drop_table()` used to support overwriting existing tables.
-   Escaping now uses the DBI generics: `sql_escape_ident()` is replaced by `dbQuoteIdentifer()` and `sql_escape_string()` is replaced by `dbQuoteString()`. (These two generics still exist inside of dplyr but are used purely for testing).

If you have provided methods any of these dbplyr generics you'll need to:

-   Implement a method for the corresponding DBI generic: this ensures that your package will work with dbplyr 2.0.0.

```{=html}
<!-- -->
```
-   Call the DBI generic from the existing dbplyr method.
    Don't delete your existing methods or your package will no longer work with dbplyr 1.0.0.

I *think* that this should cause relatively little pain because DBI support is now much better than it used to be and most backend authors provide both the dbplyr and DBI interfaces.
Even if there is some pain, I think it's relatively worthwhile pain as it helps out everyone, not just dbplyr users.

## SQL generation

There are a number of dplyr generics that basically generate SQL and then execute it.
Now rather than implementation method for a `db_*` generic from dplyr, you can implement a method for a `sql_*` generic from dbplyr.

If you have any of the following generics, extract the SQL generation code into a new `sql_` method:

    db_create_index   -> sql_index_create
    db_analyze        -> sql_table_analyze
    db_query_rows     -> sql_query_rows
    db_save_query     -> sql_query_save
    db_query_fields   -> sql_query_fields
    db_explain        -> sql_query_explain

Again, you'll need to keep your old `db_` method (so your package continues to work with dbplyr 1.0.0), but you may want to use your new `sql_` method from it to avoid code duplication.

Eventually, dbplyr will stop calling the `db_` generic and will call the `sql_` generics directly.

## Renamed generics

## Conditional method registration

To make it possible to release your package to CRAN before dbplyr 2.0.0 is out (providing the best experience for your users), you can rely on the usual `@export` roxygen tag or `S3method()` `NAMESPACE` directive because that requires importing the generic (which won't be available until after dbplyr 2.0.0 is available).

```{r}
.onLoad <- function(...) {
  if (packageVersion("dbplyr" >= "1.99")) {
    s3_register("dbplyr::sql_index_create", "backend_class")
    s3_register("dbplyr::sql_table_analyze", "backend_class")
  }
}

s3_register <- function(generic, class, method = NULL) {
  stopifnot(is.character(generic), length(generic) == 1)
  stopifnot(is.character(class), length(class) == 1)

  pieces <- strsplit(generic, "::")[[1]]
  stopifnot(length(pieces) == 2)
  package <- pieces[[1]]
  generic <- pieces[[2]]

  caller <- parent.frame()

  get_method_env <- function() {
    top <- topenv(caller)
    if (isNamespace(top)) {
      asNamespace(environmentName(top))
    } else {
      caller
    }
  }
  get_method <- function(method, env) {
    if (is.null(method)) {
      get(paste0(generic, ".", class), envir = get_method_env())
    } else {
      method
    }
  }

  method_fn <- get_method(method)
  stopifnot(is.function(method_fn))

  # Always register hook in case package is later unloaded & reloaded
  setHook(
    packageEvent(package, "onLoad"),
    function(...) {
      ns <- asNamespace(package)

      # Refresh the method, it might have been updated by `devtools::load_all()`
      method_fn <- get_method(method)

      registerS3method(generic, class, method_fn, envir = ns)
    }
  )

  # Avoid registration failures during loading (pkgload or regular)
  if (!isNamespaceLoaded(package)) {
    return(invisible())
  }

  envir <- asNamespace(package)

  # Only register if generic can be accessed
  if (exists(generic, envir)) {
    registerS3method(generic, class, method_fn, envir = envir)
  }

  invisible()
}
```

(`s3_register()` is licensed with [the unlicense](https://choosealicense.com/licenses/unlicense/) so you're free to copy and paste into your package, and there's no need to acknowledge the contribution.
Learn more about it at `?vctrs::s3_register`.)

At some point in the future, you can formally require version 2.0.0 of dbplyr in your `DESCRIPTION`, and then delete this code, switching back to your usual way of exporting methods.
